A Coq Formalization of Stable Sorts, Radix Sort, and the Burrows-Wheeler Transform

### 2019-10-11 draft_paper/draft_paper.tex
Change the title
stash

\title{Towards a Verified Bzip2}
\subtitle{A Functional Specification for Burrows--Wheeler}

@@

\begin{abstract}
Text of abstract \ldots.
\end{abstract}

@@

The Burrows--Wheeler transform (BWT) is an invertible transformation
that makes a string more amenable to compression by other methods
\cite{bw}. It accomplishes this by creating runs of repeated letters
in the produced string. Of course, a simple way to create runs is to
sort the string, but that would be useless for compression because the
original string would be unrecoverable.

To apply the transformation to a string of length $n$:
\begin{enumerate}
\item Form the \(n \times n\) matrix of all cyclic shifts of the
  original string (also referred to as the rotation matrix). Starting
  with the original string, each row is formed from the previous one
  by removing the first character and sticking it on the end.
\item Sort the rows lexicographically
\item Take the last column of the matrix. That string is the result of the transformation.
\item Note the index where the original string appears in the sorted
  rotation matrix. This piece of information is necessary to reverse
  the transformation.
\end{enumerate}
As shown in \Cref{fig:bw_ex}, applying the algorithm to the string
\var{abracadabra!} produces the new string \var{ard!rcaaabb} and the
index \var{3}. The transformation tends to produce runs like the three
\var{a}s or the two \var{b}s in \var{ard!rcaaabb}, and it is this
feature that makes the resulting string easier to compress. Although
it is not obvious, the transformed string, together with the index,
provide enough information to recover the the original string.

@@

Bird and Mu \cite{birdmu,pearls} analyze the Burrows--Wheeler algorithm
in a functional setting. They express the algorithm as the composition
of many smaller functions, and then derive an expression for the
inverse transform from the specification that it inverts the forwards
transform. In addition to providing an intuitive derivation of the
inverse transform, this provides the outline of a proof of for the
inverse BWT.

Our main contribution is a formalization in the Coq proof assistant of
the BWT, together with a machine-checked proof that our implementation
is correct. Specifically, we implement the functions $\var{bwl} :
\var{list}~A \to \var{list}~A$ and $\var{bwn} : \var{list}~A \to
\mathbb{N}$ that produce the string and index part of the BWT
respectively, and the function $\var{unbwt} : \mathbb{N} \to
\var{list}~A \to \var{list}~A$ that reverses the transform. We prove the
following theorem:
\begin{restatable*}[unbwt\_correct]{theorem}{unbwtcorrect}
  \label{thm:unbwt_correct}
  For all lists $l$,
  \begin{equation*}
    \var{unbwt}~(\var{bwn}~l)~(\var{bwl}~l) = l
  \end{equation*}
\end{restatable*}
Our implementation and proof are based on Bird et~al.'s derivations.

The implementation is useful in its own right as a functional
specification for a potential verified implementation of a program
that relies on Burrows--Wheeler, such as bzip2\cite{appel-func-spec}.
But more generally, this provides evidence that Bird's approach of
program calculation, in which programs are derived from their
specifications via algebraic manipulations, is useful for proving the
correctness of algorithms in a machine-checked setting. Bird has
applied this approach to greedy algorithms, the Boyer-Moore algorithm,
the Knuth-Morris-Pratt algorithm, Sudoku solvers, arithmetic coding,
the Schorr-Waite algorithm, the Johnson-Trotter algorithm, and many
more problems \cite{pearls}.

As a part of the proof of \Cref{thm:unbwt_correct}, we proved two
other noteworthy results that have not, to our knowledge, been proved
before in Coq. We implemented a least-significant-digit (LSD)
radixsort and proved its correctness,
\begin{restatable*}[radixsort\_correct]{theorem}{radixsortcorrect}
  \label{thm:radixsort_correct}%
  For all matrices $m$ with $n$ columns,
  \begin{equation*}
    \var{Sorted}~(\var{radixsort}~m~n) \land \var{StablePerm}~(\var{radixsort}~m~n)~m
  \end{equation*}
\end{restatable*}
and we proved that any two stable sorts are interchangeable,
\begin{restatable*}[StableSort\_unique]{theorem}{stablesortunique}
  \label{thm:stablesort_unique}
  For all functions $f, g: \var{list}~A \to \var{list}~A$,
  \begin{equation*}
    \var{StableSort}~f \implies \var{StableSort}~g \implies
    (\forall l,f~l = g~l)
  \end{equation*}
\end{restatable*}
This in turn required developing suitable definitions for
\var{Sorted}, \var{Permutation}, and \var{StablePerm}.

The rest of this paper is organized as follows. \Cref{sec:bwt} details
the history of the BWT and its applications. \Cref{sec:ord} defines
what we need to know about a type $A$ to sort a $\var{list}~A$ as a
part of BWT. \Cref{sec:forwards_BWT} implements the forwards BWT in
Coq. \Cref{sec:inverse_BWT} implements the inverse BWT and proves
\Cref{thm:unbwt_correct}. \Cref{sec:sorted_perm_stable} provides
multiple definitions of \var{Sorted}, \var{Permutation}, and
\var{StablePermutation}, and proves equivalence between them.
\Cref{sec:unique} proves \Cref{thm:stablesort_unique}. \Cref{sec:ord}
defines lexicographic orderings, need to sort the rotation matrix.
\Cref{sec:sort} radixsort and proves \Cref{thm:radixsort_correct}.
\var{sec:opt} explains the optimization that is most commonly applied
to BWT that relates our implementation to the standard version of the
algorithm.

Every definition and theorem in this paper corresponds directly to a
definition or theorem in our Coq sources, which can be found at
\url{https://github.com/jbaum98/verified_bzip2/tree/2019-05-06}.
Definitions are copied verbatim. Theorems are rephrased for clarity,
with the name of the corresponding Coq theorem given in parentheses.
The only exceptions are \var{sec:opt} and places where we give
hypothetical definitions to provide an explanation.

\section{History and applications of the BWT}
\label{sec:bwt}

The Burrows--Wheeler transform was originally introduced as a pass in a
compression algorithm, and continues to be used as a part of the
popular, open-source bzip2 compression software \cite{tsai_2016}.

Here we describe a standard compression algorithm built around BWT.
First the input is transformed with BWT, which tends to group together
characters. Intuitively, the last column of the rotation matrix
contains the characters that directly preced the characters in the
first column of the rotation matrix. Sorting lexicographically, which
completely sorts the first column. If there is a correlation between a
character and the character preceding it, bringing together identical
characters in the first column will tend to bring together identical
characters in the last column.

The next pass is a Move-to-Front (MTF) encoding, in which a character
is represented by the number of different characters encountered since
the last occurrence of this character. This creates many runs of zeroes
from the runs in the output of BWT. On English texts, often 50-60\% of
the values produced by the MTF pass are zeroes \cite{fenwick2007}.
Often a Run-Length Encoding (RLE) pass is then used to encode only the
runs of zeroes with their lengths, using a bijective base-2 encoding
\cite{bw-analysis, tsai_2016}. Finally, some sort of zeroth-order
coder like a Huffman coder is used.

BWT is also at the heart of algorithms that exploit the relationship
between the rotation matrix in BWT and the suffix array data structure
in order to search for patterns in a compressed text
\cite{ferragina_index}. These techniques have been applied to enable
fast matching of DNA subsequences \cite{dna}.

\section{Orders and preorders}
\label{sec:ord}

Although it is commonly demonstrated on strings, the BWT can be
applied to any list whose elements are comparable. To make that
precise, if we have a type $A$ and a relation on that type $\le$, we say
that this forms a preorder on $A$ when $\le$ is reflexive and transitive, that is
$x \le x$ for all $x:A$, and
\begin{equation*}
  x \le y \implies y \le z \implies x \le z
\end{equation*}
for all $x,y,z:A$.

We say that we have a \textit{total} preorder if for any two elements
$x$ and $y$, $x \le y \lor y \le x$.

We say that we have a \textit{decidable} preorder if we have a
decision procedure for determining whether $x \le y$ or $x \not\le y$.

We represent a total, decidable preorder on type with the
\var{Preord} typeclass in Coq:
\begin{lstlisting}
Class Preord (A : Type) :=
  { le : A -> A -> Prop;
    le_trans : forall x y z, le x y -> le y z -> le x z;
    le_total : forall x y, le x y \/ le y x;
    le_dec : forall x y, {le x y} + {~le x y};
  }.
\end{lstlisting}
We do not explicitly require reflexivity because it follows from
totality. Given a \var{Preord} instance for a type $A$, we will be
able to sort $\var{list}~A$ as shown in \cref{sec:sort}.

From $\le$, we can define
\begin{align*}
x \ge y \coloneqq y \le x && x < y \coloneqq \lnot (y \le x) && x > y \coloneqq \lnot (x \le y)
\end{align*}

We can also define
\begin{equation*}
  x \equiv y \coloneqq x \le y \land y \le x
\end{equation*}
$\equiv$ is a
decidable equivalence relation, which means it is reflexive,
transitive, and symmetric, and we have a decision procedure for it.

The builtin class \var{EqDec} is used to describe types with decidable
equivalence relations, so for any \var{Preord} instance we have the
\var{EqDec} instance \var{Preord\_EqDec}. In Coq, the notation
\var{=\hspace{0pt}=\hspace{0pt}=} is used for the equivalence relation of an \var{EqDec}
instance and \var{=\hspace{0pt}=} for its decision procedure.

We say that we have a total, decidable \textit{order} when we have
that for all $x$ and $y$,
\begin{equation*}
  x \equiv y \implies x = y
\end{equation*}
that is, the induced equivalence relation of the preorder ($\equiv$)
implies Leibniz equality ($=$). We use the \var{Ord} class in Coq
for this:
\begin{lstlisting}
Class Ord (A : Type) `{Preord A} :=
  { eqv_eq : forall x y, eqv x y -> x = y }.
\end{lstlisting}
For the BWT, we require an \var{Ord} instance for $A$, but
elsewhere, such as in sorting, \var{Preord} will suffice.

\section{The forwards BWT}
\label{sec:forwards_BWT}

First, we define the left and right rotations of a list. \var{lrot}
moves the first element of a list to the end.
\begin{lstlisting}
Definition lrot (l: list A) : list A :=
  match l with
  | [] => []
  | hd :: tl => tl ++ [hd]
  end.
\end{lstlisting}
\var{rrot} moves the last element of a list to the front, and
relies on \var{init}, which drops the last element of a list.
\begin{lstlisting}
Fixpoint init l : list A :=
  match l with
  | [] | [_] => []
  | hd :: tl => hd :: init tl
  end.
\end{lstlisting}
\begin{lstlisting}
Definition rrot (l: list A) : list A :=
  match l with
  | [] => []
  | hd :: tl => last l hd :: init l
  end.
\end{lstlisting}
Note that in order to ensure that $\var{last}$ is total, it takes
an extra argument to return in the case of an empty list.

Now we produce the rotation matrix by repeatedly applying \var{lrot}.
We will use \var{iter}, which takes a function $f : A \to A$, a number
$n$, and an initial value $z$, and returns the list of intermediate
values obtained by applying $f$ to $z$ $n$ times. Or,
\begin{equation}
  \var{iter}~f~n~z = [z,~f~z,~f~(f~z),~\ldots,~f^{n-1} z]
  \label{eq:iter}
\end{equation}
where $f^n$ denotes repeated function application.
\begin{lstlisting}
Fixpoint iter (f : A -> A) (n : nat) (z : A)
  : list A :=
  match n with
  | O   => []
  | S m => z :: iter f m (f z)
  end.
\end{lstlisting}

We can state \Cref{eq:iter} as a theorem by defining the
function \var{rep} for repeated function application.
\begin{lstlisting}
Fixpoint rep (f : A -> A) (n : nat) (z : A) : A :=
  match n with
  | O => z
  | S m => f (rep f m z)
  end.
\end{lstlisting}
In proofs, we will still use the notation $f^i z$ for
$\var{rep}~f~i~z$. We will also use that notation $l_i$ for the $i$th
element of a list. In Coq, that is given by $\var{nth}~i~l~d$, where
$d:A$ is a default value in case $i \ge \var{length}~l$.

\begin{theorem}[iter\_nth]
  For all $f$, $n$, $z$, and $i < n$,
  \begin{equation*}
    (\var{iter}~f~n~z)_i = f^i~z
  \end{equation*}
\end{theorem}
\begin{proof}
  The theorem follows by induction on $i$. The inductive case is
  \begin{align*}
       (\var{iter}~f~(n+1)~z)_{i+1}
    &= (z :: \var{iter}~f~n~(f~z))_{i+1} \\
    &= (\var{iter}~f~n~(f~z))_i \\
    &= f^i~(f~z) \\
    &= f^{i+1}~z \qedhere
  \end{align*}
\end{proof}

\var{bwl} produces the transformed string by taking the last column of
the sorted rotation matrix. \var{lexsort} sorts the rotation matrix
using insertion sort (\cref{appendix:insertion_sort}) and the induced
lexicographic ordering(\cref{sec:lex_ord}). By
\Cref{thm:stablesort_ord} though, \var{lexsort} can be implemented
using any sorting algorithm.
\begin{lstlisting}
Definition bwl (l : list A) : list A :=
  match l with
  | [] => []
  | hd :: _ =>
    List.map (fun x => last x hd) (lexsort (rots l))
  end.
\end{lstlisting}

\var{bwn} returns the index in the sorted rotation matrix where the
original string appears. It relies on \var{findIndex}, which is
defined with respect to some decidable equivalence relation.
\begin{lstlisting}
Fixpoint findIndex (x : A) (ls : list A) : nat :=
  match ls with
  | [] => 0
  | hd :: tl =>
    if x == hd then 0
    else S (findIndex x tl)
  end.
\end{lstlisting}

When we use \var{findIndex} in \var{bwn}, we are using the
induced equivalence relation from the \var{Ord} instance, so it
finds the row that is Leibniz equal to $l$.
\begin{lstlisting}
Definition bwn (l : list A) : nat :=
  findIndex l (lexsort (rots l)).
\end{lstlisting}

Given these definitions, we can quickly prove that \var{bwn}
satisfies its specification:
\begin{restatable*}[bwn\_correct]{theorem}{bwncorrect}
  For every non-empty list $xs$,
  \begin{equation*}
    (\var{lexsort}~(\var{rots}~xs))_{(\var{bwn}~xs)} = xs
  \end{equation*}
\end{restatable*}

First we prove a few lemmas:
\begin{lemma}[findIndex\_correct]
  For all elements $x$ and lists $l$,
  \begin{equation*}
    (\exists y \in l,~x \equiv y) \implies l_{(\var{findIndex}~x~xs)} \equiv x
  \end{equation*}
\end{lemma}
\begin{proof}
  This follows by induction on $l$: At every step, either the
  current element is equivalent to $x$ and we return $0$, or there
  must be such a $y$ in the rest of the list.
\end{proof}

\begin{lemma}[orig\_in\_sorted\_rots]
  For every non-empty list $l$,
  \begin{equation*}
    \exists y \in \var{lexsort}~(\var{rots}~l),~x = y
  \end{equation*}
\end{lemma}
\begin{proof}
  Clearly $l$ is in $\var{rots}~l$ at index 0, and because
  \var{lexsort} produces a permutation of its input, $l$ must be
  in $\var{lexsort}~(\var{rots}~l)$ as well.
\end{proof}

\bwncorrect
\begin{proof}
  This follows from \var{findIndex\_correct} if we can show that
  $\exists y \in xs,~x = y$, which we have by \var{orig\_in\_sorted\_rots}.
\end{proof}

\section{The inverse BWT}
\label{sec:inverse_BWT}

More precisely, suppose we had a function function \var{recreate}
such that
\begin{equation}
  \label{eq:recreate}
  \var{recreate}~(\var{bwl}~l) = \var{lexsort}~(\var{rots}~l)
\end{equation}
Then, we could define \var{unbwt} as
\begin{lstlisting}
Definition unbwt (i : nat) (l : list A) : list A :=
  nth i (recreate l) l.
\end{lstlisting}
and \Cref{eq:unbwt} would follow directly from the \var{bwn\_correct}:
\begin{align*}
     \var{unbwt}~(\var{bwn}~l)~(\var{bwl}~l)
  &= (\var{recreate}~(\var{bwl}~l))_{(\var{bwn})~l}\\
  &= (\var{lexsort}~(\var{rots}~l))_{(\var{bwn})~l} \\
  &= l
\end{align*}

How, then, do we implement \var{recreate} such that \Cref{eq:recreate}
holds? A key insight is that because any two methods of sorting the
rotation matrix must produce the same output, we can pretend that
\var{lexsort} is implemented using an sorting algorithm we find
convenient (\Cref{thm:sort_ord_unique}). Specifically, if we examine
how an LSD radixsort sorts the rotation matrix
(\Cref{fig:radixsort_rots}) we notice an interesting pattern. Each
column of the rotation matrix is a permutation of the original string,
so when we sort a on particular column, we always produce the same
result in that column. As we move right-to-left, the column containing
the sorted input string moves with us until it reaches the left edge
of the matrix.

@@

This suggests that we can, as we had hoped, recreate the matrix from a
single column (\Cref{fig:recreate}). But why specifically the last
column?

@@

The reason is that the last column is the solution to the fixpoint equation
\begin{equation*}
  \var{hdsort}~(\var{prefix\_shift}~c~m) = m
\end{equation*}
where \var{hdsort} sorts on the first column of a matrix,
$\var{prefix\_shift}~c$ adds $c$ as the first column of a matrix,
dropping the last column, and $m$ is the rotation matrix. This means
as we repeatedly prefix $l$ and sort, we preserve the ``rest of the
matrix'', even the parts that we have yet to reconstruct.

@@

We implement \var{recreate} as follows, introducing a parameter $j$ so
that we can recursively recreate the first $j$ columns of the matrix,
one column at a time. Here $l$ is meant to be the output of the BWT,
the last column of the sorted rotation matrix.
\begin{lstlisting}
Fixpoint recreate (j : nat) (l : list A)
  : list (list A) :=
  match j with
  | O    => map (const []) l
  | S j' => hdsort (prepend_col l (recreate j' l))
  end.
\end{lstlisting}
This definition relies on \var{prepend\_col} and \var{zipWith}, defined below.
\begin{lstlisting}
Definition prepend_col
  : list A -> list (list A) -> list (list A)
  := zipWith cons.

>>>>

\title[Coq Formalization of Stable Sorts, Radix Sort, and BWT]{A Coq
  Formalization of Stable Sorts, Radix Sort, and the Burrows-Wheeler
  Transform}

@@

\begin{abstract}
Stable sorts are an important and fundamental type of sorting
algorithm. Some algorithms, such as radix sorts, rely upon the
stability of a sorting subroutine for their correctness. In other
cases, stable sorts are useful for reasoning about algorithms, often
because any two stable sorts are interchangeable. One such algorithm
is the Burrows--Wheeler transform (BWT), which has applications in
compression, full-text indexing and pattern searching.

There has not been much work on formally verifying stable sorts, or
rather, the stability of stable sorts. While Bird
et~al.\cite{birdmu,pearls} provide an elegant derivation and proof for
the BWT, formalizing it in Coq reveals that it relies on a few lemmas
that are not trivial to prove, including the correctness of
least-significant digit (LSD) radix sort. In this paper, we develop
more fully the theory of stable sorts, and use this theory to prove
the correctness of LSD radix sort and thereby the BWT.
\end{abstract}

@@

Stable sorts are an important and fundamental type of sorting
algorithm. A sort is stable if it preserves the relative order of
equal elements. For example, if we are stably sorting the list
$[\var{banana},\ \allowbreak\var{armpit},\ \allowbreak\var{carrot},\ \allowbreak\var{apple}]$,
we have to produce
$[\var{armpit},\ \allowbreak\var{apple},\ \allowbreak\var{banana},\ \allowbreak\var{carrot}]$
and not
$[\var{apple},\ \allowbreak\var{armpit},\ \allowbreak\var{banana},\ \allowbreak\var{carrot}]$.
Even though both are sorted on the first character, because
\var{armpit} preceded \var{apple} in the original lists, it must also
precede \var{apple} in the stably sorted list.

Stability can be directly useful in certain contexts, such as when
sorting a table by a single column. However, stability is also crucial
to the correctness of certain general-purpose algorithms. One such
algorithm, or set of algorithms, is radix sort. Radix sorts are used
to sort data lexicographically, and can be more efficient than
comparison-based sorting methods, both in theory and in practice.
Radix sorts are especially well-suited to sorting strings
\cite{McIlroy93}. Many radix sorts rely on a some stable sorting
subroutine.

Stable sorts are also important for reasoning about sorting algorithms
more broadly. This is because any two stable sorting algorithms are
equivalent, in the sense that they will produce the same outputs for a
given input. Crucially, this allows us to reason using simple
algorithms and replace them with more efficient algorithms later on.
We make use of this technique in order to reason about LSD radix sort,
as well as the Burrows-Wheeler transform (BWT).

The Burrows--Wheeler transform is an invertible transformation that
makes a string more amenable to compression by other methods
\cite{bw}. Notably, the BWT is the core pass in the popular,
open-source bzip2 compression software \cite{tsai_2016}. In addition,
the BWT has been applied to full-text indexing and pattern searching,
for example in genomics \cite{ferragina_index, dna}.

There has not been much work on formally verifying stable sorts, or
rather, the stability of stable sorts. Some work has been done on
verifying the stability of mergesort
\cite{leroy_2018,Sternagel2013,Leino2015,deGouw2014}, but we do not
know of any formally verified implementation of a radix sort that
includes stability.

In addition, to our knowledge no one has produced a formally verified
implementation of the BWT, by which we mean an implementation with a
proof that the inverse-transform inverts the forwards-transform. Bird
et~al.\cite{alg-of-prog,birdmu} provide an excellent derivation of the
BWT and its inverse that we use as the basis of our implementation and
proof. However, it leaves a few crucial lemmas unproven, including the
correctness of LSD radix sort.

In this paper, we develop more fully the theory of stable sorts: we
give multiple definitions for permutations and stable permutations and
prove them equivalent. This allows us to choose whichever definition
makes the proof easiest when we are reasoning about stability. We also
proved that any two sorts are interchangeable.
\begin{restatable*}[StableSort\_unique]{theorem}{stablesortunique}
  \label{thm:stablesort_unique}
  For all functions $f, g: \var{list}~A \to \var{list}~A$,
  \begin{equation*}
    \var{StableSort}~f \implies \var{StableSort}~g \implies
    (\forall l,f~l = g~l)
  \end{equation*}
\end{restatable*}

Using \cref{thm:stablesort_unique}, we prove the correctness of an LSD
radix sort that relies on insertion sort to sort each column. This
choice was crucial to the proof: parametrizing radix sort over any
stable sort made it difficult to reason about, but reasoning about
insertion sort using induction is very elegant.
\cref{thm:stablesort_unique} allows us to recover the generality of
parametrizing over all stable sorts, because insertion sort is
equivalent to any other stable sort.

Finally, we use the correctness of radix sort to reason about the BWT.
Specifically, we implement the functions $\var{bwl} : \var{list}~A \to
\var{list}~A$ and $\var{bwn} : \var{list}~A \to \mathbb{N}$ that produce
the string and index part of the BWT respectively, and the function
$\var{unbwt} : \mathbb{N} \to \var{list}~A \to \var{list}~A$ that reverses
the transform. We then prove the following theorem:
\begin{restatable*}[unbwt\_correct]{theorem}{unbwtcorrect}
  \label{thm:unbwt_correct}
  For all lists $l$,
  \begin{equation*}
    \var{unbwt}~(\var{bwn}~l)~(\var{bwl}~l) = l
  \end{equation*}
\end{restatable*}

The implementation is useful in its own right as a functional
specification for a potential verified implementation of a program
that relies on Burrows--Wheeler, such as bzip2. But more generally,
this provides evidence that Bird's approach of program calculation, in
which programs are derived from their specifications via algebraic
manipulations, is useful for proving the correctness of algorithms in
a machine-checked setting. Bird has applied this approach to greedy
algorithms, the Boyer-Moore algorithm, the Knuth-Morris-Pratt
algorithm, Sudoku solvers, arithmetic coding, the Schorr-Waite
algorithm, the Johnson-Trotter algorithm, and many more problems
\cite{pearls}.

Every definition and theorem in this paper corresponds directly to a
definition or theorem in our Coq sources, which we provide as
supplementary materials. Definitions are copied verbatim, modulo
whitespace. Theorems are rephrased for clarity, with the name of the
corresponding Coq theorem given in parentheses. The only exceptions
are \cref{sec:opt} and places where we give hypothetical definitions
to drive the explanation.

\section{Main ideas}
\label{sec:main_ideas}

\subsection{Stable sorting}
\label{subsec:stable_sorting}

Whenver we sort a list, we do so with respect to some ordering on its
elements. For example, we could sort a list of students by first name,
last name, birthday, etc. More precisely, for sorting we need a total,
decideable preorder: a \textit{preorder} is a relation $\le$ that is
reflexive and transitive, it is \textit{total} if we can compare any
two elements, and it is \textit{decidable} if we have a procedure that
takes any two elements $x$ and $y$ and tells us if $x \le y$ or $x \ge y$.

Given a total decideable preorder, we can specify what contract a
function~$f$ must fulfill to be considered a sorting function.
Intuitively, it must produce a list that is sorted, and that contains
all of the same elements as the original list.
\begin{definition}[Sort]
  A function $f : \var{list}~A \to \var{list}~A$ is a \var{Sort} when
  for all inputs $l$,
  \begin{equation*}
    \var{Sorted}~(f~l) \land \var{Permutation}~(f~l)~l
  \end{equation*}
\end{definition}

Now we have to define \var{Sorted} and \var{Permutation}
appropriately. For both properties, and for other properties we will
discuss later, there are multiple reasonable definitions. We will
state all the definitions we can think of, and prove equivalence
between them. This increases our confidence that we have defined
things correctly, and allows us to choose whichever definition we find
most convenient for a given proof.

We can define \var{Sorted} in two main ways: %TODO cite coq book and knuth
\begin{enumerate}
\item A list $l$ is sorted when for all
  indices $i$ and $j$,
\begin{equation*}
  i \le j \implies l_i \le l_j
\end{equation*}
\item We can inductively define when a list $l$ is sorted, by saying
  that the empty list is sorted, and that given a sorted lists $t$, we
  may conclude that $h :: t$ is also sorted if $h$ is less~than or
  equal~to every element in $t$.
\end{enumerate}

Similarly, we can define \var{Permutation} in three main ways:
\begin{enumerate}
\item Two lists $l$ and $l'$ are permutations of one another if every
  $x$ occurs an equal number of times in $l$ and $l'$.
\item Two lists $l$ and $l'$ with lengths $n$ are permutations of one
  another when there exists a permutation function~$p$ that maps the
  indices of $l$ to indices of $l'$, so that for every index~$i$
  \begin{equation*}
      l_i = l'_{p(i)}
  \end{equation*}
  $p$ must be a bijection so that every index of $l$ is mapped to by
  exactly one index of $l'$.
\item We can define inductively when two lists are permutations of one another:
  \begin{itemize}
  \item As a base case, two empty lists are permutations of one
    another.
  \item If two lists $l$ and $l'$ are permutations of one another,
    then we can conclude that $x :: l$ and $x :: l'$ are permutations
    of one another.
  \item As a second base case, we can conclude that for any $x$, $y$,
    and $l$, that $x :: y :: l$ is a permutation of $y :: x :: l$.
  \item If $l_1$ is a permutation of $l_2$ and $l_2$ is a permutation
    of $l_3$, then $l_1$ is a permutation of $l_3$.
  \end{itemize}
\end{enumerate}

Now that we've defined what a sort is, we can start to consider stable
sorts. Stable sorts must preserve the relative orders of
``equivalent'' elements. This equivalence comes from the ordering we
are using: any total decidable preorder also give us a decideable
equivalence relation $\equiv$, where $x \equiv y$ when $x \le y$ and $x \ge y$.
\begin{definition}[StableSort]
  A function $f : \var{list}~A \to \var{list}~A$ is a \var{StableSort} when
  for all inputs $l$,
  \begin{equation*}
    \var{Sorted}~(f~l) \land \var{StablePerm}~(f~l)~l
  \end{equation*}
\end{definition}

Stable sorts still must produce sorted outputs, but the output cannot
be just any permutation of the input. So we need a stricter notion of
a permutation that respects equivalence, which we will call a stable
permutation.

To define stable permutations, we can tweak the each
definitions of permutation mentioned above:
\begin{enumerate}
\item Two lists $l$ and $l'$ are stable permutations of one another if
  for every $x$,
  \begin{equation*}
    \var{filter}~(\lambda y. y \equiv x)~l = \var{filter}~(\lambda y. y \equiv x)~l'
  \end{equation*}
  This is a stricter version of the first definition of permutation
  above, because that can be rephrased as $\var{filter}~(\lambda y. y = x)~l
  = \var{filter}~(\lambda y. y = x)~l'$.
\item For two lists $l$ and $l'$ of length $n$ to be stable
  permutations of one another, there must exists a \textit{stable}
  permutation function~$p$ that relates them. That means that $p$
  satisfies an additional stability property: for all indices $i < j$
  where $l_{i} \equiv l_{j}$, we must have $p(i) < p(j)$.
\item We can modify the third inference rule, which allows swapping
  $x$ and $y$, so that you can only apply it when $x \nequiv y$: the new
  rule says that for any $x$, $y$, and $l$ such that $x \nequiv y$ , you can
  conclude that $x :: y :: l$ is a permutation of $y :: x :: l$.
\end{enumerate}

From these definitions, we can see that stable sorts are restricted in
which permutation they produce. In fact, there is only one stable
permutation that is also sorted, so any two stable sorts must produce
the same results for every input, and are thus interchangeable.
\stablesortunique

This can be generalized a little further: even if a sorting algorithm
itself is not a stable sorting algorithm, if the ordering we are using
doesn't allow us to distinguish between equivalent elements, then we
will not be able to tell if equivalent elements are being swapped.
More precisely, notice that it isn't always true that $x \equiv y$ implies
$x = y$. For example, if we are sorting strings by their first
character, then $\text{\texttt{apple}} \equiv \text{\texttt{armpit}}$ under
this ordering, but clearly $\text{\texttt{apple}} \neq
\text{\texttt{armpit}}$. When we do have that the equivalence ($\equiv$)
implies Leibniz equality ($=$), we say that we have an \textit{order},
as we have when we are sorting strings using all of their characters.
Under an order, as opposed to a preorder, we can prove that every
permutation is a stable permutation and every sort is a stable sort,
which implies that every sort is equivalent to every other sort.
\begin{theorem}[all\_perm\_stable]
  If we have an equivalence that implies Leibniz equality, then for
  all lists $l$ and $l'$
  \begin{equation*}
      \var{Permutation}~l~l' \implies \var{StablePerm}~l~l'
  \end{equation*}
\end{theorem}
\begin{corollary}[Sort\_Ord\_unique]
  \label{thm:sort_ord_unique}
  When sorting using an order, given any two function $f, g:
  \var{list}~A \to \var{list}~A$,
  \begin{equation*}
    \var{Sort}~f \implies \var{Sort}~g \implies
    (\forall l, f~l = g~l)
  \end{equation*}
\end{corollary}

\subsection{LSD radix sort}
\label{sec:radix_sort}

LSD radix sort is a simple radix sort that is used to
lexicographically sort a list of strings of equal length. LSD radix
sort operates by stably sorting on each column, from the end of the
string to the front (\cref{fig:lsd_radixsort}). The algorithm works
because lexicographic order says that if two strings are equal up to a
column, then the tie is broken according to the rest of the string.
When we stably sort on a column, ties are broken in favor or the
existing order, which was established by sorted on the following
columns.

\subsection{The Burrows-Wheeler transform}
\label{subsec:bwt}

The Burrows--Wheeler transform was originally introduced as a pass in a
compression algorithm \cite{bw}, and continues to be used as a part of
the popular, open-source bzip2 compression software \cite{tsai_2016}.
Here we describe a standard compression algorithm built around BWT.
\begin{enumerate}
\item Apply the BWT, which tends to group create runs of identical characters.
\item Apply Move-to-Front (MTF) encoding, in which a character is
  represented by the number of different characters encountered since
  the last occurrence of this character. This creates many runs of
  zeroes from the runs in the output of BWT. On English texts, often
  50-60\% of the values produced by the MTF pass are zeroes
  \cite{fenwick2007}.
\item Apply Run-Length Encoding (RLE) to encode only the runs of
  zeroes with their lengths, using a bijective base-2 encoding
  \cite{bw-analysis, tsai_2016}.
\item Compress with some zeroth-order coder like a Huffman coder.
\end{enumerate}

The BWT is also at the heart of algorithms that exploit the
relationship between the rotation matrix in the BWT and the suffix
array data structure in order to search for patterns in a compressed
text \cite{ferragina_index}. These techniques have been applied to
enable fast matching of DNA subsequences \cite{dna}.

To apply the BWT to a string of length $n$:
\begin{enumerate}
\item Form the \(n \times n\) matrix of all cyclic shifts of the
  original string (referred to as the rotation matrix). Starting
  with the original string, each row is formed from the previous one
  by removing the first character and sticking it on the end.
\item Sort the rows lexicographically
\item Take the last column of the matrix. That string is the result of
  the transformation.
\item Note the index where the original string appears in the sorted
  rotation matrix. This piece of information is necessary to reverse
  the transformation.
\end{enumerate}
As shown in \cref{fig:bw_ex}, applying the algorithm to the string
\var{abracadabra!} produces the new string \var{ard!rcaaabb} and the
index \var{3}.

@@

The BWT tends to produce runs like the three \var{a}s or the two
\var{b}s in \var{ard!rcaaabb}, and it is this feature that makes the
resulting string easier to compress. It creates those runs by
exploiting the correlation between adjacent characters, which means
that by knowing one character, one can predict the character that
follows it with some success. Sorting the rotation matrix
lexicographically brings together rows that start with the same
characters. But, the beginning and end of each row were adjacent in
the original string, so this will also tend to bring together rows
that end with the same characters.

\subsection{Inverting the BWT}
\label{subsec:bwt}

Although it is not obvious, the transformed string and the index
provide enough information to recover the the original string.
Conceptually, we can find the original string by using the transformed
string to reconstruct the entire sorted rotation matrix, and then
using the index to find the row containing the original string. This,
though, relies on our ability to reconstruct the sorted rotation
matrix from just its last column.

A key insight is that because of \cref{thm:sort_ord_unique}, we can
pretend that during the forwards transform, the rotation matrix is
being sorted using any sorting algorithm we find convenient.
Specifically, if we examine how an LSD radixsort sorts the rotation
matrix (\cref{fig:radixsort_rots}) we notice an interesting pattern.
Each column of the rotation matrix is a permutation of the original
string, so when we sort a on particular column, we always produce the
same result. As we move right-to-left, the column containing the
sorted input string moves with us until it reaches the left edge of
the matrix.

@@

This suggests that we can recreate the matrix from the last column by
sorting it to produce the first column, then repeatedly prepending the
last column to the growing matrix and stably sorting by it
(\cref{fig:recreate}. As the matrix grows from right to left, we are
mimicing the process of LSD radix sort in \cref{fig:radixsort_rots},
in that the sorted portions to the right of the shaded column are
identical.

@@

The choice of the last column is crucial, because as the ``original
string'' in the stable column sort, it acts as the tie-breaker. The
last column is the solution to the fixpoint equation
\begin{equation*}
  \var{hdsort}~(\var{prefix\_shift}~c~m) = m
\end{equation*}
where \var{hdsort} sorts on the first column of a matrix,
$\var{prefix\_shift}~c$ adds $c$ as the first column of a matrix,
dropping the last column, and $m$ is the rotation matrix. This means
as we repeatedly prefix $l$ and sort, we preserve the ``rest of the
matrix'', even the parts that we have yet to reconstruct.

@@

\section{Implementing the forwards BWT}
\label{sec:forwards_BWT}

First, we define the left and right rotations of a list. \var{lrot}
moves the first element of a list to the end.
\begin{lstlisting}
Definition lrot (l: list A) : list A :=
  match l with
  | [] => []
  | hd :: tl => tl ++ [hd]
  end.
\end{lstlisting}
\var{rrot} moves the last element of a list to the front, and
relies on \var{init}, which drops the last element of a list.
\begin{lstlisting}
Definition rrot (l: list A) : list A :=
  match l with
  | [] => []
  | hd :: tl => last l hd :: init l
  end.
\end{lstlisting}
Note that in order to ensure that $\var{last}$ is total, it takes
an extra argument to return in the case of an empty list.
